---
title: 'The Vitamin C intake of the Guinea Pig: data analysis'
author: "Ilya Maltsev"
date: "March 13, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The Vitamin C intake of the Guinea Pig can be observed by the growth 
of the odontoblasts of the incisor tooth 
([original paper](http://jn.nutrition.org/content/33/5/491.full.pdf)). 
[The observed dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) contains measured length of odontoblasts for one of three dose 
levels by one of two delivery methods. The project focuses on comparing 
length versus dose and supplement. 

It is proved that the bigger is dose the larger is length for any supplement.
'Orange juice' supplement results in larger length for the doses 0.5 and 1, 
no significant difference is found for the dose 2.

## Data processing

Requred libraries:
```{r, load libraries, results='hide'}
library(datasets)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```

Load the data:
```{r, load the data}
data("ToothGrowth")
ToothGrowth <- tbl_df(ToothGrowth)
```

### Basic summary

```{r, missed values}
table(is.na.data.frame(ToothGrowth))
```
There are no missed values in the data.

Basic properties of the data:
```{r, basic summary}
summary(ToothGrowth)
```

Number of available observations of 'orange juice' delivery method is `r table(ToothGrowth$supp)["OJ"]` and `r table(ToothGrowth$supp)["VC"]` 
for 'vitamin c' delivery method

### Exploratory analysis

```{r, sample distribution of length, echo=FALSE}
ggplot(ToothGrowth, aes(x = len)) + 
    geom_histogram(bins = 20, alpha = 0.2, colour = "black") + 
    facet_grid(dose ~ supp) +
    ggtitle("Length distribution for given dose and supplement")
```

It is obvious from the plot that the dose affects the length, the bigger 
is dose the larger is length. The distribution of lengths given dose and 
supplement does not seem to match any known distribution.

Assume the lengths follow to normal distribution (null hypothesis). Number 
of observations is small so the hypothesis can be verified by [Shapiro–Wilk test](https://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test). Try to reject 
the null hypothesis given an $\alpha$ level of 0.05.

```{r, shapiro-wilk test}
alpha <- 0.05
# if reject = FALSE then we failed to reject
ToothGrowth %>% group_by(supp, dose) %>%
    summarise(pvalue = shapiro.test(len)$p.value, reject = pvalue <= alpha)
```

We failed to reject the hypothesis for all combinations of supplement and dose.
It means we can assume that the lengths distribution is normal.

## Dosage and supplement effect

Let the null hypothesis is some (supp2, dose2) combination has greater effect 
than some other combination (supp1, dose1). Let the probability of rejection 
($\alpha$, Type 1 error rate) is 0.05. We would reject the null hypothesis 
if right bound of confidence interval is greater than zero or if p-value is 
greater than $\alpha$.

```{r, dosage test}
alpha <- 0.05
data_frame(dose1 = c(0.5,  1,    0.5,  1,    0.5,  1,    2),
           dose2 = c(1,    2,    1,    2,    0.5,  1,    2),
           supp1 = c("OJ", "OJ", "VC", "VC", "VC", "VC", "VC"),
           supp2 = c("OJ", "OJ", "VC", "VC", "OJ", "OJ", "OJ")) %>%
    rowwise() %>% # evaluate by row
    mutate(htest = list(t.test(
        y = (ToothGrowth %>% filter(supp == supp2) %>% filter(dose == dose2))$len,
        x = (ToothGrowth %>% filter(supp == supp1) %>% filter(dose == dose1))$len,
        alternative = "less", paired = FALSE, 
        var.equal = FALSE, conf.level = 1 - alpha))) %>% # run test
    mutate(pvalue = htest$p.value) %>% # extract p-value
    mutate(left = htest[[4]][1]) %>% # extract left bound of conf.int
    mutate(right = htest[[4]][2]) %>% # extract right bound of conf.int
    mutate(reject = pvalue > alpha | right > 0) %>% # make a decision
    select(-htest)
```

The test shows that increasing of the dose given a supplement results in 
increasing of the length in all cases. Length effect of 'orange juice' 
supplement is bigger than 'vitamin c' one for the doses 0.5 and 1. 
The test cannot distinguish length effect of the supplements for dose equal 
to 2 for given level of significance.

## Conclusion

The distribution of length given supplement and dose can be assumed normal with
the significance level $\alpha = 0.5$. This is proved by Shapiro-Wilk test.

T confidence interval test shows the bigger is dose the larger is length and 
'orange juice' supplement results in larger length for the doses 0.5 and 1.
Significance level is $\alpha = 0.5$. No significant difference is found 
for the dose 2.
